\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-12 analysis/paper/paper.Rmd

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- FALSE
eval_cs2 <- TRUE
eval_cs3 <- FALSE
eval_cs4 <- FALSE
```

@@

In the next case study, we show how case-base sampling can also be used in the context of a competing risk analysis. For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the main author's website, and it is also available as part of the \pkg{casebase} package.

@@

We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [-@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital. Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R}. Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest. \textcolor{red}{How come we lose individuals during imputation?} The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual. A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}. The data was downloaded from the main author's website, and it is also available as part of the \pkg{casebase} package.

@@

Hospital death as a covariate was removed, as this is directly informative of death. The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase. To visualize the incidence density, we can refer to the post-imputation population-time plot in Figure \ref{fig:ptPostI}.

```{r supportSetUp, echo = FALSE, eval = eval_cs3}
completeData=casebase::support
completeData<-completeData[,-c(3)]
#keep one individual out, at random, for which we will develop an 
#absolute risk curve.
sam <- sample(1:nrow(completeData), 1)
#Set workingData to the remaining individuals in the data
#workingData=completeData[-c(sam),]
workingCompleteData <- as.matrix(sparse.model.matrix(death~ . - d.time, data = completeData))[,-c(1)]
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData <- workingCompleteData[sam,]
x <- workingCompleteData[-c(sam),]
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y <- data.matrix(completeData[-c(sam), c(34, 4)])
```

```{r poptimePreImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPreI} Population-time plot before imputation. Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
popTime_pre <- popTime(as.data.frame(support),
                       time = "d.time", event = "death")
plot(popTime_pre)
```

```{r poptimePostImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPostI} Population-time plot before imputation. Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
popTime_post <- popTime(as.data.frame(completeData),
                        time = "d.time", event = "death")
plot(popTime_post)
```

Our main objective is to estimate the cumulative incidence of death for a given covariate profile. Given th number of covariates available, we will do this using regularized estimation. First, we fit the hazard function; for the sake of simplicity, we opted for a linear term for time. As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}. Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables. Here, we used lasso penalization by setting \code{alpha} to 1.

```{r CBhazard, echo = TRUE, eval = eval_cs3}
cb.Model <- fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", 
                                event = "death", formula_time = ~ d.time, 
                                alpha = 1, ratio = 100)
```

Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function.

```{r CBabsolute, echo = TRUE, eval = eval_cs3}

casebaseAbsolute <- absoluteRisk(cb.Model,
                                 time = seq(1, max(completeData$d.time), 0.5),
                                 newdata = t(newData),
                                 s = "lambda.1se", method = "numerical")
```

We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation. However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}. To circumvent this issue and estimate the baseline hazard, we followed the following steps. First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object. Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}. Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.

```{r coxHazAbsolute, echo = TRUE, eval = eval_cs3}
# Create u and xCox, which will be used when fitting Cox with glmnet.
u <- survival::Surv(time = as.numeric(y[,2]), 
                    event = as.factor(y[,1]))
xCox <- as.matrix(sparse.model.matrix(death~ . - d.time,
                                      data = completeData))[-c(sam),]
# hazard for cox using glmnet
coxglmFit <- glmnet::cv.glmnet(x = xCox, y = u,
                               family = "cox", alpha = 1)
# convergence demonstrated in plot
# plot(coxglmFit)
#taking the coefficient estimates for later use
nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
# creating a new dataset that only contains the covariates chosen through glmnet.
cleanCoxData <- as.data.frame(cbind(as.numeric(y[,2]),
                                    as.numeric(y[,1]),
                                    xCox[, nonzero_covariate_cox$X1]))
# fitting a cox model using regular estimation, however we will not keep it.
# this is used more as an object place holder.
coxFit <- survival::coxph(Surv(time = V1, event = V2) ~ ., 
                          data = cleanCoxData)
# The coefficients of this object will be replaced 
# with the estimates from coxglmFit.
coxFit$coefficients <- nonzero_coef_cox@x
# Fitting CI curve for cox+glmnet
newDataCox <- newData[nonzero_covariate_cox$X1 - 1]

abCoxFit <- survival::survfit(coxFit, newdata = as.data.frame(t(newDataCox)),
                              time = seq(0, max(completeData$d.time), 1),
                              type = "breslow")
```

@@

```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
km <- survival::Surv(time = completeData$d.time, 
                     event = completeData$death)
abKm <- survival::survfit(km ~ 1,
                          type = 'kaplan-meier')
```

Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier.

```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
plot(casebaseAbsolute, type='l', col = "black", lwd = 2, 
     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
lines(abCoxFit, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
legend("bottom", 
       legend = c("casebase", "Cox+glmnet", "KM curve"), 
       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
```

>>>>

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- FALSE
eval_cs2 <- FALSE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
```

@@

In the next case study, we show how case-base sampling can also be used in the context of a competing risk analysis. For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the first author's website, and it is also available as part of the \pkg{casebase} package.

@@

In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves. In the next two case studies, we explore more advanced features.

For the third case study, we will look at regularized estimation for the hazard function. This can be performed using a combination of case-base sampling and regularized logistic regression. To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital. Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R} [@mice]. Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 individuals. \textcolor{red}{How come we lose individuals during imputation?} Of these, 67.5% experienced the event of interest. <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}. The original data was obtained from the website of the Department of Biostatistics at Vanderbilt University. The imputed dataset is available as part of the \pkg{casebase} package.

@@

<!--Hospital death as a covariate was removed, as this is directly informative of death. The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.--> As before, the first step is to visualize the incidence density. The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}. \textcolor{red}{This population-time plot is strange. The red dots never seem to go above 200.}

```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for tje SUPPORT data."}
data(support)
support_popTime <- popTime(support,
                           time = "d.time", event = "death")
plot(support_popTime)
```

Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death and cannot be used as a predictor. Then we select one observation at random. We will estimate the cumulative incidence curve for this individual. The other observations will be used to fit the hazard function.

```{r supportSetUp, echo = TRUE, eval = eval_cs3}
support <- dplyr::select(support, -hospdead)

# Randomly select one observation
indiv <- sample(1:nrow(support), size = 1)
```

Our main objective is to estimate the cumulative incidence of death for a given covariate profile. Given the number of covariates available, we will do this using regularized estimation. As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}. Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables.

```{r support-design, eval = eval_cs3}
# Create design matrix without the time variable
support_matrix <- model.matrix(death ~ . - d.time, 
                               data = support)
# Remove intercept term
support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]

# Create test and train dataset
support_test <- support_matrix[indiv,,drop = FALSE]

support_train <- support_matrix[-indiv,]
y <- as.matrix(support[-indiv, c("death", "d.time")])
```

We are now ready to estimate the hazard function. We use lasso penalization; this is achieved by setting \code{alpha} to 1. Also, note that the argument \code{formula_time} can be used to specify the function form for time; in our example, we model time linearly.

```{r CBhazard, echo = TRUE, eval = eval_cs3}
support_cb <- fitSmoothHazard.fit(support_train, y, 
                                  formula_time = ~ d.time, 
                                  family = "glmnet", 
                                  event = "death", time = "d.time", 
                                  alpha = 1, ratio = 100)
```

Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function. Since \code{glmnet} fitted the penalized logistic regression for a range of penalty terms, we need to specify which one we want to use for our prediction. We select \code{s = "lambda.1se"}, as recommended.

```{r CBabsolute, echo = TRUE, eval = eval_cs3}
time_points <- seq(1, max(support$d.time), by = 0.5)
support_risk <- absoluteRisk(support_cb,
                             time = time_points,
                             newdata = support_test,
                             s = "lambda.1se", method = "numerical")
```

We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation.

```{r coxHaz, echo = TRUE, eval = eval_cs3}
# Create survival object for glmnet
surv_glmnet <- survival::Surv(time = y[, "d.time"], 
                              event = y[, "death"])

support_cox <- glmnet::cv.glmnet(x = support_train, 
                                 y = surv_glmnet,
                                 family = "cox", alpha = 1)
```

However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}. To circumvent this issue and get an estimate of the baseline hazard, we followed the following steps. First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object. Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}. Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.

```{r coxAbsolute, echo = TRUE, eval = eval_cs3}
# Identifying selected covariates
nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
                                  s = "lambda.1se")
nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
# Creating a new dataset that only contains the covariates selected by glmnet
support_cox_data <- as.data.frame(cbind(y,
                                        support_train[, nonzero_covariates_cox$X1]))
# Fitting a cox model
support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
                          data = support_cox_data)
# Replace the estimate with those obtained by glmnet
support_cox$coefficients <- nonzero_coef_cox@x

# Fitting CI curve for cox+glmnet
support_cox_risk <- survival::survfit(support_cox,
                                      newdata = as.data.frame(support_test),
                                      time = time_points,
                                      type = "breslow")
```

@@

```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
km <- survival::Surv(time = support$d.time, 
                     event = support$death)
abKm <- survival::survfit(km ~ 1,
                          type = 'kaplan-meier')
```

Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier. \textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}

```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue."}
plot(support_risk, type='l', col = "black", lwd = 2, 
     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
lines(support_cox_risk, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
legend("bottom", 
       legend = c("casebase", "Cox+glmnet", "KM curve"), 
       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
```
